Broom: Safe Region-Based Memory Management for Dataflow Programs

### 2015-11-09 paper/introduction.tex
Motivating examples added.

Consider the example shown in Fig.~\ref{fig:motivating-eg}.
\TODO{Replace by more complex example where data-structure is updated
over multiple invocations of method and finally transferred.}
This code represents the logic for a streaming query operator (also known
as a continuous query). The operator receives a continuous stream of
input messages, which are batched into time-windows associated with a logical
timestamp. Messages may be delivered out-of-order and multiple messages with
the same timestamp may be delivered. A timing-message, periodically indicates that
no more input messages with a timestamp less than a given value $t$ will be subsequently
delivered. At this point, the operator can create the output message corresponding
to time $t$ and deliver it to its successors.

@@

This example is an instance of a general pattern, where a producer creates a data-structure
and passes it to a consumer.
In a system where most of the computation takes this form and these data-structures are
very large, as is the case with many streaming big-data analysis systems,
the garbage collection overheads become significant.
% as the collector scans these large data-structures
However, these data-structures have a well-defined lifetime, which can be exploited to
achieve better performance 
% using alternative memory management techniques.

>>>>

Consider the example shown in Fig.~\ref{fig:motivating-eg}.
This code represents the logic for a streaming query operator (also
known as a continuous query). The operator receives a continuous
stream of input messages, which are processed via a user-defined
function, and batched into time-windows associated with a logical
timestamp. Messages may be delivered out-of-order and multiple
messages with the same timestamp may be delivered. A timing-message,
periodically indicates that no more input messages with a timestamp
less than or equalt to a given value $t$ will be subsequently
delivered. At this point, the operator can create the output message
corresponding to time $t$ and deliver it to its successors.

@@

This example is an instance of a general pattern, where a producer
creates a data-structure and passes it to a consumer.  In a system
where most of the computation takes this form, and these
data-structures are very large, as is the case with many streaming
big-data analysis systems, the garbage collection overheads become
significant.
% as the collector scans these large data-structures
However, these data-structures have a well-defined lifetime, which can
be exploited to achieve better performance 
% using alternative memory management techniques.

### 2015-11-09 paper/related.tex
Motivating examples added.

Tofte and Talpin in~\cite{tofte93,tofte94,tofte97} introduce the
concept of a region type system to statically ensure the safety of
region-based memory management in ML. Following their seminal work,
static type systems for safe region-based memory management have been
extensively studied in the context of various languages and problem
settings~\cite{cyclone02, cyclone04, yates99, MIT03, DPJ09, HMN01,
WW01, rust}. Our work differs from the existing proposals in a number
of ways. Firstly, our problem setting includes lexically scoped stack
regions and dynamic transferable regions (both programmer-managed) in
context of an object-oriented programming language equipped with
higher-order functions. Second, we adopt a two-pronged approach to
memory safety that relies on a combination of a simple static type
system and lightweight runtime checks. The type system reduces the
problem of ensuring safety of all memory accesses in a program into a
problem of enforcing a state-transition discipline on memory accesses
over a small subset of objects in the program. The later can be done
efficiently at runtime. This lets us guarantee safety in presence of
dynamic memory allocations and deallocations without requiring either
restrictive static mechanisms (e.g., linear types and unique pointers)
or expensive runtime mechanisms (e.g., garbage collection and
reference counting). Lastly, our region type system comes equipped
with full type inference that completely eliminates the need to write
region annotations on types to convince the type checker that the
program is safe.

>>>>

Tofte and Talpin in~\cite{tofte93,tofte94,tofte97} introduce the
concept of a region type system to statically ensure the safety of
region-based memory management in ML. Following their seminal work,
static type systems for safe region-based memory management have been
extensively studied in the context of various languages and problem
settings~\cite{cyclone02, cyclone04, yates99, MIT03, DPJ09, HMN01,
WW01, rust}. Our work differs from the existing proposals in a number
of ways. Firstly, our problem setting includes lexically scoped stack
regions and dynamic transferable regions (both programmer-managed) in
context of an object-oriented programming language equipped with
higher-order functions. Second, we adopt a two-pronged approach to
memory safety that relies on a combination of a simple static type
discipline and lightweight runtime checks. 
% The type system reduces the problem of ensuring safety of all memory
% accesses in a program into a problem of enforcing a state-transition
% discipline on memory accesses over a small subset of objects in the
% program. The later can be done efficiently at runtime. This lets us
% guarantee safety in presence of
In particular, our approach requires neither restrictive static
mechanisms (e.g., linear types and unique pointers) nor expensive
runtime mechanisms (e.g., garbage collection and reference counting)
in order to guarantee safety. Lastly, our region type system comes
equipped with full type inference that completely eliminates the need
to write region annotations on types to convince the type checker that
the program is safe.
